Parsing Postfix log files

### 2009-03-11 doc/logparser-implementation.tex
Rearranged sections; the content hasn't changed, just the ordering.
Corrections for the assumptions and database sections of the implementation chapter.

XXX FIGURE OUT WHEN TO USE ACTIONS AND WHEN TO USE ALGORITHM, ETC\@.

XXX THIS MIGHT BE USEFUL SOMEWHERE\@:

XXX SHOULD I SAY ``cleaned up'' EVERYWHERE OR NOT\@?

Whereas the rules are quite simple and each rule is completely independent
of the other rules, the algorithm is significantly more complicated and
highly internally interdependent.  The algorithm deals with all the
complications of parsing, the eccentricities and oddities of Postfix log
files, presenting the resulting data in a normalised, simple to use
representation.  The algorithm's task is to follow the journey each mail
takes through Postfix, combining the data extracted by rules into a
coherent whole, saving it in a useful and consistent form, and performing
housekeeping duties.

@@

    \item The log files are whole and complete: nothing has been removed,
        either deliberately or accidentally (e.g.\ log file rotation gone
        awry, file system filling up, logging system unable to cope with
        the volume of log messages).  On a well run system it is extremely
        unlikely that any of these problems will arise, though it is of
        course possible, particularly when suffering a deluge of spam or a
        mail loop.

    \item Postfix logs enough information to make it possible to accurately
        reconstruct the actions it has taken.  There are several heuristics
        used when parsing; see
        \sectionref{identifying-bounce-notifications},
        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
        logging after cleanup} for details.

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  There are
advantages to treating Postfix as a black box during parser development:

@@

    \item The parser is developed using real world log files rather than
        the idealised log files someone would naturally envisage reading
        the source code.

    \item The source code cannot accurately communicate the variety of
        orderings in which log lines are written to the log file, as
        process scheduling independently interferes with logging and other
        processing.

    \item The parser acts as a second source of information about Postfix's
        operation, using information gathered from empirical evidence.  A
        separate project could compare the empirical knowledge inherent in
        the parsing algorithm with the documentation and source code of
        Postfix to see how closely the two agree.

@@

\section{Database Schema}
\label{database schema}

The database is an integral part of the parser presented here: it stores
the rules and the data gleaned by applying those rules to Postfix log
files.  Understanding the database schema is important in understanding the
actions of the parser, and essential to developing further applications
which utilise the data gathered.

The database schema can be conceptually divided in two: the rules used to
parse log files, and the data saved from the parsing of log files.  Rules
have the fields required to parse the log lines, extract data to be saved,
and specify the action to be executed; they also have several fields that
aid the user in understanding the meaning of the log lines parsed by each
rule.  The rules are described in detail in \sectionref{rules in
implementation} but the fields are covered in \sectionref{rule attributes}.

The data saved from parsing the log files is also divided into two tables
as described below: connections and results.  The connections table
contains a row for every mail accepted and every connection where there was
a rejection; the individual fields will be described in
\sectionref{connections table}.  There will be at least one row in the
results table for each row in the connections table; the fields will be
covered in detail in \sectionref{results table}.

An important but easily overlooked benefit of storing the rules in the
database is the link between rules and results: if more information is
required when examining a result, the rule that produced the result is
available for inspection: each result references the rule that created it.
There is no ambiguity about which rule resulted in a particular result,
eliminating one potential source of confusion.

@@

The database populated by this program provides a simple interface to
Postfix log files.  Although the interface is a database schema rather than
a set of functions provided by a library, it performs the same function as
any other \acronym{API}: it provides a stable interface, allowing code on
either side of the interface to be changed without adverse effects, as long
as the interface is maintained.  The parser can be improved to handle
additional complications, support earlier or later releases of Postfix,
bugs can be fixed or limitations removed.  Programs that use the database
can range from the simple examples in \sectionref{motivation} to far more
complex data mining tools.

@@

    \item Facilities are provided to access databases from most programming
        languages, allowing a developer to access the data gathered using
        their preferred programming language, rather than being restricted
        to the language the parser is written in.  It is often possible to
        write an interface layer allowing code written in one language to
        be used in another language, but this greatly increases the effort
        required to use the parser.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide a program, of varying complexity and
        sophistication, that can be used for ad hoc queries with minimal
        investment of time.

@@

            \item If the database supports it, access can be granted on a
                fine-grained basis, e.g.\ allowing the finance department
                to produce invoices, the helpdesk to run limited queries as
                part of dealing with support calls, and the administrators
                to have full access to the data.

@@

\begin{verbatim}
CREATE TRIGGER ON INSERT INTO results
    WHERE sender = 'boss@example.com'
        AND postfix_action = 'REJECTED'
    SEND PANIC EMAIL TO 'postmaster@example.com';
\end{verbatim}

            \item Other tables can be added to the database, e.g.\ to cache
                historical, summary, or computed data.

@@

    \item \acronym{SQL} is reasonably standard and many people will already
        be familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn: a good
        introduction to \acronym{SQL} can be found at
        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
        Although every vendor implements a different dialect of
        \acronym{SQL}, the basics are the same everywhere (analogous to the
        overall similarities and minor differences amongst Irish English,
        British English, American English, and Australian English).
        Depending on the database in use there may be tools available that
        reduce or remove the requirement to know \acronym{SQL}; several are
        available for \gls{SQLite}, the default database used by the parser
        implementation.

@@

@@

Rules are discussed in detail in \sectionref{rules in implementation}, but
the structure of the rules table is covered here.  Rules are created by the
user, not the parser, and will not be modified by the parser (except for
the hits and hits\_total fields).  Rules parse the individual log lines,
extracting data to be saved in the connections and results tables, and
specifying the action to take for that log line.

@@

\begin{eqlist}

@@

    \item [restriction\_name] The restriction that caused the mail to be
        rejected.  Only applicable to rules that have an action of
        \texttt{REJECTION}, other rules should have an empty string.

@@

    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
        log lines the rule applies to.  This avoids needlessly trying rules
        that will not match the log line, or worse, might match
        unintentionally.  Rules whose program is \texttt{*} will be tried
        against any log lines that are not parsed by program specific
        rules.

    \item [regex] The regex to match the log line against.  The regex will
        have several keywords expanded when the rules are loaded: this
        simplifies reading and writing rules; avoids needless repetition of
        complex regex components; allows the components to be corrected
        and/or improved in one location; enables automatic extraction and
        saving of data; and makes each regex largely self-documenting.  For
        efficiency the keywords are expanded and every rule's regex is
        compiled before attempting to parse the log file --- otherwise each
        regex would be recompiled each time it was used, resulting in a
        large, data dependent slowdown.  Efficiency concerns are discussed
        in \sectionref{parser efficiency}, with caching compiled regexes
        covered in \sectionref{Caching compiled regexes}.

    \item [connection\_data] Sometimes rules need to save data that is not
        present in the log line: e.g.\ setting \texttt{client\_ip} when a
        mail is being delivered to another server.  The format is:
        \newline{} \tab{} \texttt{ client\_hostname = localhost,}
        \newline{} \tab{} \tab{} \texttt{client\_ip
        = 127.0.0.1} \newline{} i.e.\ semi-colon or comma separated
        assignment statements, with the variable name on the left and the
        value on the right hand side.\footnote{Commas and semi-colons
        cannot be escaped and thus cannot be used.  This feature is
        intended for use with small amounts of data rather than large
        amounts in any one rule, so dealing with escape sequences was
        deemed unnecessary.}  Any field in the connections table can be set
        in this way.

@@

    \item [action] The action that will be invoked when this rule matches a
        log line; a full list of actions and the parameters they are
        invoked with can be found in \sectionref{actions in detail in
        implementation}.

    \item [hits] This counter is maintained for every rule and incremented
        each time the rule successfully matches.  At the start of each run
        the parser sorts the rules in descending order of hits, and at the
        end of the run it updates every rule's hits field in the database.
        Assuming that the distribution of log lines is reasonably
        consistent across log files, rules matching more commonly occurring
        log lines will be tried before rules matching less commonly
        occurring log lines, reducing the parser's execution time.  Rule
        ordering for efficiency is discussed in \sectionref{rule ordering
        for efficiency}.

@@

    \item [priority] This is the user-configurable companion to hits: when
        the list of rules is sorted, priority overrides hits.  This allows
        more specific rules to take precedence over more general rules
        (described in \sectionref{overlapping rules in architecture}).

%    \item [cluster\_group] A reference to the \texttt{cluster\_group}
%        table.  That table is used by the Decision Tree algorithm
%        described in a separate document.

@@

@@

Every log line a row in the results table, unless the
\texttt{postfix\_action} field of the rule which parsed it was INFO or
IGNORE\@.  Each row in the results table is associated with a single
connection, and there may be many results per connection.

\begin{eqlist}

@@

    \item [rule\_id] The id of the entry in the rules table that matched
        the log line and created this result.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \acronym{SMTP} command
        --- a mechanism that is quite useful for testing new restrictions.
        This field will be 1 if the log line parsed was a warning rather
        than a real rejection, or 0 for a real rejection.  This field
        should be ignored if the result is not a rejection, i.e.\ the
        action field of the associated rule is not \texttt{REJECTION}.

    \item [smtp\_code] The \acronym{SMTP} code associated with the log
        line.  In general an \acronym{SMTP} code is only present in a log
        line for a rejection or final delivery; results initially missing
        an \acronym{SMTP} code will duplicate the \acronym{SMTP} code of
        other results in that connection.  Some final delivery log lines do
        not contain an \acronym{SMTP} code: in those cases the code is
        specified by the rule, based on the success or failure represented
        by the log line.

@@

    \item [sender] The sender's email address.  Because multiple mails may
        be delivered during one connection, there may be different sender
        addresses in the results for one connection; however there should
        not be different sender addresses in the results for one mail.
        Mails sent over the same connection can be distinguished by their
        queueid.

@@

    \item [delay] How long the mail was delayed for while it was being
        delivered.  This will only be present for delivered mails.

    \item [delays] More details information about how long the mail was
        delayed for while it was being delivered  This will only be present
        for delivered mails.

>>>>

XXX FIGURE OUT WHEN TO USE ACTIONS AND WHEN TO USE ALGORITHM, PARSER,
ETC\@.

The parser deals with all the complications of parsing, the eccentricities
and oddities of Postfix log files, presenting the resulting data in a
normalised, simple to use representation.  The parser's task is to follow
the journey each mail takes through Postfix, combining the data extracted
by rules into a coherent whole, saving it in a useful and consistent form,
and performing housekeeping duties.

@@

    \item The log files are whole and complete: nothing has been removed,
        either deliberately or accidentally (e.g.\ log file rotation gone
        awry, file system filling up, logging system unable to cope with
        the volume of log messages).  On a well run system it is extremely
        unlikely that any of these problems will arise, though it is of
        course possible, particularly when suffering from a deluge of spam
        or a mail loop.

    \item Postfix logs enough information to make it possible to accurately
        reconstruct the actions it has taken.  There are several heuristics
        used when parsing; see
        \sectionref{identifying-bounce-notifications},
        \sectionref{aborted-delivery-attempts}, and \sectionref{pickup
        logging after cleanup} for details.

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  XXX REFER TO
BLACK BOX PAPER\@?  There are some advantages to treating Postfix as a
black box during parser development:

@@

    \item The parser is developed using real world log files rather than
        the idealised log files someone would naturally envisage reading
        the source code.  The source code cannot accurately communicate the
        variety of orderings in which log lines are written to the log
        file, as process scheduling independently interferes with logging
        and other processing.

    \item The parser acts as a second source of information about Postfix's
        operation, using information gathered from empirical evidence.  A
        separate project could compare the empirical knowledge inherent in
        the parsing algorithm with Postfix's documentation and source code
        to see how closely the two agree.

@@

\section{Database}

\label{database}

The database is an integral part of the parser presented here: it stores
the rules and the data gleaned by using those rules to parse Postfix log
files.  Understanding the database schema is important in understanding the
actions of the parser, and essential to developing further applications
which utilise the data gathered; \sectionref{database as API} describes
how the database schema functions as an \acronym{API}.

The database schema can be conceptually divided in two: the rules used to
recognise log lines, and the data saved from the parsing of log files.
Rules have the fields required to recognise log lines, extract data to be
saved, and specify the action to be invoked; they also have several fields
that aid the user in understanding the meaning of the log lines recognised
by each rule.  The rules are described in detail in \sectionref{rules in
implementation}, but the fields are covered with the rest of the database
schema in this section.

The data saved from parsing the log files is also divided into two tables:
connections and results.  The connections table contains an entry for every
mail accepted and every connection where there was a rejection; the
individual fields will be described in \sectionref{connections table}.
There will be at least one entry in the results table for each entry in the
connections table; the result table's fields will be covered in detail in
\sectionref{results table}.  A diagram of the database schema is provided
in \sectionref{Diagram of the Database Schema} before the in-depth
descriptions of each table.

An important but easily overlooked benefit of storing the rules in the
database is the link between rules and results: if more information is
required when examining a result, the rule that produced the result is
available for inspection because each result references the rule that
created it.  There is no ambiguity about which rule resulted in a
particular result, eliminating one potential source of confusion.

@@

The database populated by \parsername{} provides a simple interface to
Postfix log files.  Although the interface is a database schema rather than
a set of functions in a library, it has the same benefits as any other
\acronym{API}: it provides a stable interface, allowing code on either side
of the interface to be changed without adverse effects, as long as the
interface is adhered to.  Programs that use the database can range from the
simple examples in \sectionref{motivation} to far more complex data mining
tools.

@@

    \item Most programming languages have facilities for database access,
        allowing a developer to write programs that use the gathered data
        in their preferred programming language, rather than being
        restricted to the language the parser is written in.

    \item Databases provide complex querying and sorting functionality for
        the user without requiring large amounts of programming.  All
        databases provide a program, of varying complexity and
        sophistication, that can be used for ad hoc queries with minimal
        investment of time.

@@

            \item Some databases support granting access on a fine-grained
                basis, e.g.\ allowing the finance department to produce
                invoices, the helpdesk to run limited queries as part of
                dealing with support calls, and the administrators to have
                full access to the data.

@@

\begin{verbatim}
CREATE TRIGGER ON INSERT INTO results
    WHERE sender = 'boss@example.com'
        AND action = 'DELIVERY_REJECTED'
    SEND PANIC EMAIL TO 'postmaster@example.com';
\end{verbatim}

            \item Other tables can be added to the database, to cache
                historical, summary, or computed data.

@@

    \item \acronym{SQL} is reasonably standard and many people will already
        be familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn, e.g.\
        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
        Although every vendor implements a different dialect of
        \acronym{SQL}, the basics are the same everywhere, analogous to the
        overall similarities and minor differences amongst Irish English,
        British English, American English, and Australian English.
        Depending on the database in use there may be tools available that
        reduce or remove the requirement to know \acronym{SQL}; several are
        available for \gls{SQLite}, the default database used by
        \parsername{}.

@@

\subsection{Diagram of the Database Schema}

\label{Diagram of the Database Schema}

@@

Rules are discussed in detail in \sectionref{rules in implementation}, but
the structure of the rules table is covered here with the other tables in
the database.  Rules are created by the user, not the parser, and will not
be modified by the parser (except for the hits and hits\_total fields).
Rules classify the individual log lines, extracting data to be saved in the
connections and results tables, and specifying the action to take for that
log line.

@@

\begin{boldeqlist}

@@

    \item [restriction\_name] The restriction that caused the mail to be
        rejected.  Only applicable to rules that recognise rejection log
        lines, i.e.\ rules that have an action of
        \texttt{DELIVERY\_REJECTED}, other rules should have an empty
        string.

@@

    \item [program] The Postfix component (e.g.\ \daemon{smtpd}) whose log
        lines the rule applies to.  This avoids needlessly trying rules
        that will not match the log line, or worse, might match
        unintentionally.  Rules whose program is \texttt{*} will be tried
        against any log lines that are not parsed by program specific
        rules.  This is a simple example of the rule conditions described
        in \sectionref{attaching conditions to rules}.

    \item [regex] The regex to recognise log lines with.  The regex will
        have several keywords expanded when the rules are loaded: this
        simplifies reading and writing rules; avoids needless repetition of
        complex regex components; allows the components to be corrected
        and/or improved in one location; enables automatic extraction and
        saving of data; and makes each regex largely self-documenting.  For
        efficiency the keywords are expanded and every rule's regex is
        compiled before attempting to parse the log file; otherwise every
        regex would be recompiled each time it was used, resulting in a
        large, data dependent slowdown, as shown in \sectionref{Caching
        compiled regexes}.

    \item [connection\_data] Sometimes rules need to save data that is not
        present in the log line: e.g.\ setting \texttt{client\_ip} when a
        mail is being delivered to another server.  The format is:
        \newline{} \tab{} \texttt{ client\_hostname = localhost,}
        \newline{} \tab{} \tab{} \texttt{client\_ip = 127.0.0.1} \newline{}
        i.e.\ semi-colon or comma separated assignment statements, with the
        variable name on the left and the value on the right hand
        side.\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be included in data.  This feature is intended for use with
        small amounts of data, so dealing with escape sequences was deemed
        unnecessary.}  Any field in the connections table can be set in
        this way.

@@

    \item [action] The action that will be invoked when this rule
        recognises a log line; a full list of actions and the parameters
        they are invoked with can be found in \sectionref{actions in detail
        in implementation}.

    \item [hits] This counter is maintained for every rule and incremented
        each time the rule successfully recognises a log line.  At the
        start of each run the parser sorts the rules by how frequently each
        rule recognised log lines, and at the end of the run it updates
        every rule's hits field in the database.  Assuming that the
        distribution of log lines is reasonably consistent across log
        files, ordering rules by their recognition frequency will reduce
        the parser's execution time.  Rule ordering for efficiency is
        discussed in \sectionref{rule ordering for efficiency}.

@@

    \item [priority] This is the user-configurable companion to hits: when
        the list of rules is sorted, priority overrides hits.  This allows
        more specific rules to take precedence over more general rules
        (described in \sectionref{rules in architecture}).

\end{boldeqlist}

@@

\end{boldeqlist}

@@

XXX REWRITE THIS PARAGRAPH\@; FIGURE OUT WHAT THE HELL IS HAPPENING WITH
POSTFIX\_ACTION\@.

Every log line will have a row in the results table, unless the \newline{}
\texttt{postfix\_action} field of the rule which parsed it was INFO or
IGNORE\@.  Each row in the results table is associated with a single
connection; a single connection will have at least one result associated
with it, but will usually have several more, and may have hundreds.

\begin{boldeqlist}

@@

    \item [rule\_id] The id of the rule in the rules table that recognised
        the log line and created this result.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \acronym{SMTP} command
        --- a mechanism that is quite useful for testing new restrictions.
        This field will be set to 1 if the log line parsed was a warning rather
        than a real rejection, or to 0 for a real rejection.  This field
        should be ignored if the result is not a rejection, i.e.\ the
        action field of the associated rule is not
        \texttt{DELIVERY\_REJECTED}.

    \item [smtp\_code] The \acronym{SMTP} code associated with the log
        line.  In general an \acronym{SMTP} code is only present in a log
        line for a rejection or final delivery; results whose log line did
        not contain an \acronym{SMTP} code will duplicate the
        \acronym{SMTP} code of other results in that connection.  Some
        final delivery log lines do not contain an \acronym{SMTP} code: in
        those cases the \acronym{SMTP} code is specified by the rule's
        \texttt{result\_data} field, based on the success or failure
        represented by the log line.

@@

    \item [sender] The sender's email address.  Because multiple mails may
        be delivered during one connection, there may be different sender
        addresses in the results for one connection; however there should
        not be different sender addresses in the results for one mail.
        Mails sent over the same connection can be distinguished by their
        queueids.

@@

    \item [delay] How long the mail was delayed while it was being
        delivered.  This will only be present for delivered mails.

    \item [delays] More detailed information about how long the mail was
        delayed while it was being delivered, again only present for
        delivered mails.

### 2009-03-11 doc/logparser-introduction.tex
Corrections for the assumptions and database sections of the implementation chapter.

Chapter~\ref{Postfix Parser Implementation} discusses the parsing rules in
detail, explaining the purpose and usage of each field in a rule, referring
to an example rule and sample data it matches successfully against.  The
pros and cons of overlapping rules are considered, including techniques for
detecting unintentional overlaps.  Rule efficiency concerns are discussed,
in particular the optimisations used by the algorithm.  The section
concludes with a description of using the tools provided with the parser to
generate new rules (specifically the regex in each rule) from unparsed log
lines.  It contains the core of the paper, describing a naive parsing
algorithm and the complications initially encountered that shaped the full
algorithm.  A flow chart and a discussion of the emergent behaviour
exhibited by the algorithm accompanies a comprehensive explanation of the
different stages of the initial algorithm.  The framework that actions and
rules fit into is documented, then the actions taken during execution of
the algorithm are described, followed by the process of adding a new
action.  The section concludes with an in-depth description of the further
complications discovered, and their solutions that complete the parser.
This algorithm requires a database for storing both the rules used when
parsing and the results gleaned from parsing.  The database schema used is
described in \sectionref{database schema}, explaining in detail the tables
used for storing the data gleaned from the log files and the table that
stores the rules.

>>>>

Chapter~\ref{Postfix Parser Implementation} discusses the parsing rules in
detail, explaining the purpose and usage of each field in a rule, referring
to an example rule and sample data it matches successfully against.  The
pros and cons of overlapping rules are considered, including techniques for
detecting unintentional overlaps.  Rule efficiency concerns are discussed,
in particular the optimisations used by the algorithm.  The section
concludes with a description of using the tools provided with the parser to
generate new rules (specifically the regex in each rule) from unparsed log
lines.  It contains the core of the paper, describing a naive parsing
algorithm and the complications initially encountered that shaped the full
algorithm.  A flow chart and a discussion of the emergent behaviour
exhibited by the algorithm accompanies a comprehensive explanation of the
different stages of the initial algorithm.  The framework that actions and
rules fit into is documented, then the actions taken during execution of
the algorithm are described, followed by the process of adding a new
action.  The section concludes with an in-depth description of the further
complications discovered, and their solutions that complete the parser.
This algorithm requires a database for storing both the rules used when
parsing and the results gleaned from parsing.  The database schema used is
described in \sectionref{database}, explaining in detail the tables used
for storing the data gleaned from the log files and the table that stores
the rules.
